---
title: "Getting Started with menstrualcycleR and Phase-Aligned Cycle Time Scaling (PACTS)"
author: "Anisha Nagpal, Tory Eisenlohr-Moul"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with menstrualcycleR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  cache = FALSE
)
```

# Installation of the `menstrualcycleR` package

Currently the `menstrualcycleR` package is hosted on the Eisenlohr-Moul Lab's (CLEAR Lab: Clarifying the Endocrinology of Acute Risk) GitHub repository. 

To install it from GitHub, install and load the package `remotes`` by running: 

```{r remotes, eval = F}
install.packages("remotes")
```

```{r loading-remotes, warning = F, eval = F}
library(remotes)
```

then install and load the `menstrualcycleR` package by running: 

```{r menstrualcycleR-package, eval = F}
remotes::install_github("eisenlohrmoullab/menstrualcycleR")
```

```{r loading menstrualcycleR-package}
library(menstrualcycleR)
```

Note that the `menstrualcycleR` package depends on several other packages (mostly packages from the [`tidyverse` suite](https://www.tidyverse.org/)).

This vignette also uses `tidyverse` functions so it is recommended to install and load the `tidyverse` suite by running:


```{r tidyverse, eval=FALSE}
install.packages("tidyverse")
```


```{r loading-tidyverse, warning=FALSE, eval=FALSE}
library(tidyverse)
```

Alternatively, individual packages can be loaded independently:

```{r loading-pcks, warning=FALSE}
library(dplyr)
library(ggplot2)
library(purrr)
library(tibble)
library(tidyr)
library(stringr)
library(magrittr)
```


Additionally in this vignette, we will cover options to use the continuous cycle time measures generated by `menstrualcycleR` in nonlinear multilevel analyses using the `mgcv` and `gam.hp` packages. You can install and load these packages by running: 

```{r gamm-packages-install, eval = F}
install.packages("mgcv")
install.packages("gam.hp")
```

```{r}
library(mgcv)
library(gam.hp)
```

# Demo dataset 

To explore the `menstrualcycleR` package, we will use a demo dataset with the data (daily ratings) from 








