---
title: "Preparing Your Data for PACTS"
author: "Tory Eisenlohr-Moul"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Preparing Your Data for PACTS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

This vignette covers getting data into the shape `pacts_scaling()` expects, and
choosing which of its output variables to model. For what PACTS is and why continuous
cycle time outperforms day counting, see `vignette("menstrualcycleR-overview")` or the
[visual explainer](https://menstrualcycler.clearlabresearch.com/pacts-explainer.html).

```{r libs, message = FALSE, warning = FALSE}
library(menstrualcycleR)
library(dplyr)
```

# Input format

`pacts_scaling()` takes one long-format table with four required columns.

| Column | Contents | Notes |
|---|---|---|
| `id` | Person identifier | Any type. One person may contribute many cycles. |
| `date` | Date of the observation | Class `Date`, not character. |
| `menses` | `1` on the first day of bleeding, `0` otherwise | Not every bleeding day. See Section 3. |
| `ovtoday` | `1` on the estimated day of ovulation, `0` otherwise | Required as a column even without an ovulation biomarker. See Section 4. |

Outcome columns sit alongside these. `pacts_scaling()` does not modify them; it appends
18 new columns, of which four are the cycle-time variables covered in Section 7.

```{r shape}
head(cycledata, 4)
```

# One row per person per day

**No duplicate `id`-`date` pairs.** Where a participant submitted two diary entries on
one day, resolve them to a single row before scaling. Two rows for one date means that
day is counted twice in every downstream model.

```{r dupes}
dups <- cycledata %>%
  count(id, daterated) %>%
  filter(n > 1)
nrow(dups)
```

**Gaps need no filling.** `pacts_scaling()` fills the calendar internally so that cycle
position is defined for every date inside a cycle. It therefore returns more rows than it
was given, and the added rows carry a date and a cycle-time value but no outcome.
`cycledata` goes in with 619 rows and comes out with 744, so 125 rows (17% of the output)
have no `symptom` value. Count coverage over rows where the outcome is observed, not over
all rows.

```{r coverage}
scaled <- pacts_scaling(cycledata, id = id, date = daterated,
                        menses = menses, ovtoday = ovtoday)

c(all_rows             = nrow(scaled),
  rows_with_an_outcome = sum(!is.na(scaled$symptom)))
```

Any per-person day count or sparsity threshold should use the second number.

The `lower_cyclength_bound` / `upper_cyclength_bound` arguments default to 21 and 35 days and
gate **ovulation imputation** only. A cycle with a confirmed ovulation is scaled whatever its
length, subject to its phase lengths instead. Nagpal et al. (2025) §2.1.1 describes the bounds
as deciding which cycles are scaled at all; the package applies the carve-out named in that
section's next sentence, and does so by default rather than on request. `summary_ovulation()`
reports how many cycles fall outside 21-35 days. To exclude them, filter on
`mcyclength_complete` after scaling.

# Menses onset

`menses == 1` marks the first day of a period; every later bleeding day is `0`. Coding
all bleeding days as `1` makes the function read a four-day period as four cycle onsets,
and nothing errors.

Exclude periovulatory spotting, which is not a menses onset and will start a phantom
cycle. Reclassify post-midnight entries to the previous day before scaling, so the diary
and any sleep or wearable data agree on which day is which.

# Ovulation

`ovtoday` marks the estimated day of ovulation. Ovulation biomarkers are preferred, and the coding
convention differs by method.

| Method | Code `ovtoday = 1` on | Report |
|---|---|---|
| Urinary LH test | The day after the first positive test (LH+1) | Brand and threshold, e.g. 40 mIU/ml |
| Basal body temperature | The day after the temperature nadir | How temperature was measured |
| Daily hormone assay | See Nagpal et al. (2025) | Assay and analyte |
| None available | Pass `ovtoday = NULL` | The imputation rate (Section 7) |

With only period-tracker dates, pass `ovtoday = NULL`. The column is created for you as
all-`NA`, `pacts_scaling()` places ovulation 15 days before each next menses onset, records
it in `ovtoday_impute`, and covers those cycles in the `*_impute` cycle-time columns. A
message reports that this happened. Supplying your own column of all `0`s or `NA`s does the
same thing.

Omitting the argument altogether is an error rather than a silent fallback, so that
forgetting it when you *do* have ovulation data cannot quietly produce an all-imputed
analysis. Against 33 hormone-confirmed cycles the imputed estimate was off by a mean
absolute 0.97 days (Section 7.2).

# From a period-tracker export to the PACTS table

The common starting point is one table of period start dates and a separate table of
daily measurements.

```{r sources}
period_log <- data.frame(
  id           = c(1, 1, 1, 2, 2, 2),
  period_start = as.Date(c("2024-01-03", "2024-01-31", "2024-02-27",
                           "2024-01-08", "2024-02-10", "2024-03-09"))
)

set.seed(1)
wearable <- data.frame(
  id   = rep(c(1, 2), each = 75),
  date = c(seq(as.Date("2024-01-01"), by = "day", length.out = 75),
           seq(as.Date("2024-01-05"), by = "day", length.out = 75)),
  resting_hr = round(c(rnorm(75, 58, 2), rnorm(75, 62, 2)), 1)
)

head(period_log, 3)
head(wearable, 3)
```

A left join turns the period log into a daily flag.

```{r join}
daily <- wearable %>%
  left_join(mutate(period_log, menses = 1L),
            by = c("id", "date" = "period_start")) %>%
  mutate(
    menses  = coalesce(menses, 0L),
    ovtoday = 0L          # no ovulation biomarker in this example
  )

daily %>% filter(menses == 1) %>% head(3)
```

```{r scale-example}
daily_scaled <- pacts_scaling(daily, id = id, date = date,
                              menses = menses, ovtoday = ovtoday)

daily_scaled %>%
  select(id, date, resting_hr, cyclic_time_impute, ovtoday_impute) %>%
  filter(!is.na(cyclic_time_impute)) %>%
  head(4)
```

## Common device exports

Reshape each source into `id`, `date`, value form before running the join above.

- **Oura.** Daily sleep and readiness endpoints already give one row per day. Use the
  sleep `day` field rather than the timestamp, so nights are attributed to the day the
  person woke.
- **Apple Health.** Exports are per-sample. Aggregate to a daily value first, choosing
  explicitly between a daily mean, a nightly minimum, or a single reading. Apple Health
  also records `MenstrualFlow`, which can supply `menses` if reduced to the first day of
  each run of flow days.
- **Fitbit.** Daily summaries arrive as one row per day per metric, in separate files.
  Join them on date before joining to the period log.
- **Period-tracker apps.** Most export one row per bleeding day rather than one per
  period. Reduce to onsets first by keeping a day only when the previous calendar day
  was not also a bleeding day.

# Checks before scaling

```{r checks}
checks <- c(
  date_is_Date        = inherits(daily$date, "Date"),
  no_duplicate_rows   = !any(duplicated(daily[c("id", "date")])),
  menses_is_binary    = all(daily$menses %in% c(0, 1)),
  has_ovtoday_column  = "ovtoday" %in% names(daily),
  every_id_has_onset  = all(tapply(daily$menses, daily$id, sum) > 0)
)
checks
```

A `FALSE` on the last check usually means a person's period dates failed to join, most
often because the identifier is a factor in one table and a character in the other.

After scaling, `cycledata_check()` reports how much usable outcome data each person has
in each phase, which feeds any inclusion rule.

It imposes no threshold. How much coverage a person needs before their scaled data support an
estimate is an open question, along with power, cyclical clustering, and the reliability of
per-person smooths. See "Outstanding Questions" in
`vignette("menstrualcycleR-overview")`.

# Choosing a cycle-time variable

`pacts_scaling()` returns four cycle-time variables, crossing the anchor with whether
imputed ovulation is included.

|  | Confirmed ovulation only | Imputed ovulation included |
|---|---|---|
| **Centered on menses onset** | `cyclic_time` | `cyclic_time_impute` |
| **Centered on ovulation** | `cyclic_time_ov` | `cyclic_time_imp_ov` |

Switching anchor changes no row's inclusion. Switching between confirmed-only and
imputation-inclusive changed coverage from 358 rows to 735 in `cycledata`, and from 0 rows
to 744 in a copy of it with `ovtoday` zeroed.

## Anchor

Center on the event the outcome is tied to. Perimenstrual outcomes such as dysmenorrhea
belong on the menses-centered variables, periovulatory outcomes such as reward
sensitivity on the ovulation-centered ones. Nagpal et al. (2025) recommend fitting both
in separate models where the expected timing is unclear.

Each variable places one anchor at zero and the other at the endpoints, `-1` and `+1`.
On `cyclic_time`, menses onset is at zero and ovulation is at the wrap; on
`cyclic_time_ov`, the reverse. A feature sitting at the wrap is split across the two ends
of the plotted axis, and any summary of where it falls has to be circular rather than
linear, since an arithmetic mean of positions straddling `-1` and `+1` is meaningless.
Nagpal et al. (2025) additionally flag, as an open limitation rather than a settled
result, that centering on a hormonal transition may reduce sensitivity to change at the
endpoints, and suggest alternative centering such as the mid-follicular phase as future
work.

Fit both models when the outcome has features at both anchors, such as a perimenstrual
rise and a periovulatory dip, and read each feature from the fit that centers it.

`cyclic_time` and `cyclic_time_ov` cover the same rows as each other, and so do
`cyclic_time_impute` and `cyclic_time_imp_ov`. This held on `cycledata` and on a copy
chopped to create a left-censored opening tail and an open trailing cycle.

Having no ovulation biomarker does not change which anchor to pick, but it does put a measured event
at zero in one case and an estimated one at zero in the other. The imputed ovulation day is
off by a mean absolute 0.97 days (Section 7.2). Spread over a follicular phase averaging
14.8 days, that is 0.066 of the 0-to-1 follicular span, so a feature narrower than about
0.07 on the axis cannot be located reliably in an ovulation-centered fit. State in the
caption that zero is an estimated day.

## Confirmed or imputed ovulation

In `cycledata`, 14 of 25 cycles have a confirmed ovulation. `cyclic_time` scales 358 of
744 rows and `cyclic_time_impute` scales 735. Zeroing `ovtoday` throughout drops
`cyclic_time` to 0 rows and raises `cyclic_time_impute` to 744, because every cycle then
takes the imputed path.

Run this on your own data before choosing.

```{r coverage-by-variable}
vars <- c("cyclic_time", "cyclic_time_impute", "cyclic_time_ov", "cyclic_time_imp_ov")
sapply(vars, function(v) sum(!is.na(scaled[[v]])))
```

Model `cyclic_time` or `cyclic_time_ov` when the confirmed-only coverage is enough for
the analysis, and report the confirmation rate either way. When it is not, use the
`*_impute` variables rather than dropping the unconfirmed cycles: the alternative to an
imputed ovulation is a forward day count, which Nagpal et al. (2025) found had the highest
within-person residual variance of the five time variables they compared.

Against 33 hormone-confirmed cycles, the day −15 backward count differed from the
hormone-confirmed day of ovulation by a mean absolute 0.97 days (SD 0.88), with a mean
signed difference of 0.30 days (SD 1.29). The error grew with cycle length in the same
sample (*r* = 0.395, *p* = .023), so the estimate is least accurate in the longest cycles.
PACTS has not been validated in samples with highly variable cycle lengths, frequent
anovulation, or conditions that alter hormone dynamics.

## Sensitivity to imputed ovulation

No confirmation rate has been established below which ovulation-anchored questions
become unanswerable. The published guidance is to acknowledge the imprecision and report
the proportion of cycles in which ovulation was estimated rather than measured.

Fit the model twice whenever any cycle in the analysis used an imputed ovulation: once
on confirmed cycles only, once on the full set. Nagpal et al. (2025) report that their
confirmed-only analyses agreed with the combined dataset, which is what justified pooling
those 44 cycles. That agreement is a property of a particular sample, not of the method.

```{r sensitivity}
# The confirmed-only subset is the rows cyclic_time covers.
c(rows_confirmed_only = sum(!is.na(scaled$cyclic_time)),
  rows_incl_imputed   = sum(!is.na(scaled$cyclic_time_impute)))
```

Report both fits in the methods, including when they agree.

## Extended-phase rows

This filter is package behavior rather than published method. When a cycle's luteal or
follicular phase falls outside the plausible bounds, the `*_impute` variables still
cover it, using phase fractions computed without the upper phase-length cap. Those rows
carry a flag.

```{r extended-filter, eval = FALSE}
strict <- subset(scaled, cyclic_time_impute_extended_phase == 0)
```

In `cycledata` this flag is set on 5.4% of scaled rows. The fallback only adds coverage
and has no argument that disables it, so filter after scaling if you do not want those
rows.

# What to report

**How ovulation was determined.** `summary_ovulation()` splits cycles into
biomarker-confirmed and imputed.

```{r ovsummary}
summary_ovulation(scaled)$ovstatus_total
```

**What share of coverage came from the phase-length fallback.**

```{r extended}
mean(scaled$cyclic_time_impute_extended_phase == 1, na.rm = TRUE)
```

# References

Nagpal, A., Schmalenberger, K. M., Barone, J. C., Mulligan, E., Stumper, A., Knol, L.,
Failenschmid, J., Kiesner, J., Peters, J. R., & Eisenlohr-Moul, T. A. (2025). Studying
the menstrual cycle as a continuous variable: Implementing Phase-Aligned Cycle Time
Scaling (PACTS) with the `menstrualcycleR` package. *Psychoneuroendocrinology*, 107584.
<https://doi.org/10.1016/j.psyneuen.2025.107584>

Schmalenberger, K. M., et al. (2021). How to study the menstrual cycle: Practical tools
and recommendations. *Psychoneuroendocrinology, 123*, 104895.
<https://doi.org/10.1016/j.psyneuen.2020.104895>

----

*This vignette was drafted with Claude Code and reviewed by the package authors, who are
responsible for its content. See the README for the package's full statement on AI tool
use.*
